---
title: "OSM_microbiome_workflow_smoking_status"
output: html_document
date: "2026-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis Overview

This analysis implements an OSM microbiome workflow to characterize and compare microbial community composition between smokers and non-smokers. The objective of this workflow is to assess whether smoking status is associated with differences in microbiome diversity, structure, and taxonomic abundance.

The workflow begins with the import and preprocessing of microbiome data, including quality control, filtering, and normalization steps to ensure comparability across samples. Metadata describing participant characteristics, including smoking status, is integrated to enable group-based comparisons.

Subsequent analyses focus on:

Alpha diversity to evaluate within-sample microbial richness and evenness.

Beta diversity to examine differences in overall microbial community composition between smokers and non-smokers.

Ordination and visualization techniques to explore clustering patterns related to smoking status.

Differential abundance analyses to identify specific taxa associated with smoking behavior.

All analyses are conducted within a reproducible R Markdown framework, allowing for transparent documentation of methods, results, and interpretations. This workflow is designed to provide a comprehensive overview of how smoking may influence the oral/OSM microbiome.


## Libraries

This section clears the R environment and loads the packages required for data manipulation, visualization, ecological analysis, and microbiome-specific statistical modeling. These libraries support diversity analyses, multivariate methods, contamination correction, and differential abundance testing used throughout the OSM microbiome workflow.

```{r Library}

rm(list = ls())

# library for plotting
library(ggplot2)
# library for most analysis functions such as PCA etc
library(vegan)
# library for maaslin
#BiocManager::install("Maaslin2")
library(Maaslin2)

library(tidyr)
library(dplyr)
library(microDecon)
library(readr)

```


## Data Preparation

Metadata and taxonomic abundance data are imported and harmonized for downstream microbiome analyses. Samples are filtered to define three analysis datasets: (i) all biological samples excluding contact lens samples, (ii) samples including negative controls, and (iii) samples excluding both positive and negative controls. Taxonomic profiles are restricted to the species level, non-bacterial taxa are removed, and sample identifiers are standardized between metadata and abundance tables. Species and samples with zero abundance across all observations are excluded, and metadata are refiltered to match the final abundance matrices.


```{r Data Preparation}

# Load of data

#Read metadata
metadata <- read.csv("../INPUT/OSM_metadata_smokers_contact_complete.csv", sep = ",", header = TRUE)
metadata$treatment <- ifelse(metadata$treatment == "control", "non-smoker", metadata$treatment)

# Load the RDS file
data_bacteria <- readRDS("../INPUT/metaphlan/merged_output.rds")
data_fungi <- read_tsv("../INPUT/micop/merged_taxonomy_fungi.tsv")
data_virus <- read_tsv("../INPUT/micop/merged_taxonomy_viral.tsv")


#-------
  
#Data Preparation--------------------------------
#We prepare 4 metadata

#Metadata general 
metadata_smokers_general <- metadata[metadata$treatment!="contact lense", ]
metadata_smokers_general$sample <- gsub("_$","", metadata_smokers_general$sample )
metadata_smokers_general <-metadata_smokers_general[metadata_smokers_general$sample_name!="18_CR", ]

#Metadata smokers with negative control
metadata_smokers_with_Neg <- metadata_smokers_general[metadata_smokers_general$treatment!="positive control", ]

#Metadata without Postive and negative control without decontamination
metadata_smokers <- metadata_smokers_with_Neg[metadata_smokers_with_Neg$treatment!="negative control", ]

#Preparation of Data RDS
#Selecting only the species level
#Bacteria
data_species_bacteria <- data_bacteria$species
colnames(data_species_bacteria) <- gsub("*X","", colnames(data_species_bacteria) )
colnames(data_species_bacteria) <- gsub("_metaphlan","", colnames(data_species_bacteria) )
data_species_bacteria <- data_species_bacteria[!grepl("k__Eukaryota", data_species_bacteria$clade_name), ]

#Fungi
data_species_fungi <- data_fungi[data_fungi$RANK=="species", ]
data_species_fungi <- data_species_fungi[,-c(1:3)]
colnames(data_species_fungi) <- gsub("_micop_fungi_abundance.txt","", colnames(data_species_fungi) )
names(data_species_fungi)[names(data_species_fungi) == "TAXPATHSN"] <- "clade_name"

#Virus
data_species_virus <- data_virus[data_virus$RANK=="species", ]
data_species_virus <- data_species_virus[,-c(1:3)]
colnames(data_species_virus) <- gsub("_micop_viral_abundance.txt","", colnames(data_species_virus) )
names(data_species_virus)[names(data_species_virus) == "TAXPATHSN"] <- "clade_name"

data_species_virus <- as.data.frame(data_species_virus)
# Remove rows where clade_name contains 'Human endogenous retrovirus' or 'HERV'
data_species_virus <- data_species_virus[!grepl("Human endogenous retrovirus|HERV", data_species_virus$clade_name, ignore.case = TRUE), ]

```


```{r filtering_of_microorganisms }

source("Functions_presentation/Filtering_preNorm.R") 

filtered_bacteria <- filter_species_data(
  data_species_bacteria,
  metadata_smokers_general,
  metadata_smokers_with_Neg,
  metadata_smokers
)

filtered_fungi <- filter_species_data(
  data_species_fungi,
  metadata_smokers_general,
  metadata_smokers_with_Neg,
  metadata_smokers
)


filtered_virus <- filter_species_data(
  data_species_virus,
  metadata_smokers_general,
  metadata_smokers_with_Neg,
  metadata_smokers
)

metadata_smokers_virus <- filtered_virus$metadata_smokers

```



## Decontamination with microDecon

This step applies microDecon to remove potential contaminant taxa using negative control samples. Species identified as contaminants are filtered from the abundance table when sufficient taxa are present, after which negative control samples, zero-abundance species, and empty samples are removed. Metadata are refiltered to match the decontaminated dataset, producing a cleaned species table and corresponding metadata for downstream analyses.

```{r MicroDecon_step }

#Bacteria

source("Functions_presentation/Microdecon_method.R") 

result_bacteria <- run_microdecon(filtered_bacteria$data_with_Neg, filtered_bacteria$metadata_smokers)

data_species_filtered_decontam_bacteria <- result_bacteria$data
metadata_smokers_decontam_bacteria <- result_bacteria$metadata
removed_clades_bacteria <- result_bacteria$removed_otus

#Fungi

result_fungi <- run_microdecon(filtered_fungi$data_with_Neg, filtered_fungi$metadata_smokers)

data_species_filtered_decontam_fungi <- result_fungi$data
metadata_smokers_decontam_fungi <- result_fungi$metadata
removed_clades_fungi <- result_fungi$removed_otus


#Virus

result_virus <- run_microdecon(filtered_virus$data_with_Neg, filtered_virus$metadata_smokers)

data_species_filtered_decontam_virus <- result_virus$data
metadata_smokers_decontam_virus <- result_virus$metadata
removed_clades_virus <- result_virus$removed_otus


```

## Re-normalization

Sequencing statistics are integrated with the decontaminated species abundance table to adjust for library size differences across samples. When sufficient taxa are present, read counts are recalibrated and abundances are renormalized; otherwise, the original abundances are retained. Pseudo-count–based absolute abundance estimates are then computed for downstream statistical analyses.



```{r Re-normalization}
# Normalization functions and stats table (loaded once)
source("Functions_presentation/Normalization_and_Stats_reads.R") 
stats_table <- read.csv("../../INPUT/stats_table.csv", header = TRUE)
source("Functions_presentation/ReNormalization.R") 


##BACTERIA
# Prepare stats
stats_subset_table_bacteria <- prepare_stats_table(
  stats_table = stats_table,
  metadata_general = metadata_smokers_general,
  metadata_target = metadata_smokers_decontam_bacteria
)

stats_subset_bacteria <- stats_subset_table_bacteria$stats_subset
stats_subset_general_bacteria <- stats_subset_table_bacteria$stats_subset_general

# Renormalization
renorm_bacteria <- renormalize_if_needed(
  data = data_species_filtered_decontam_bacteria,
  stats_subset = stats_subset_bacteria,
  read_column_suffix = "metaphlan"
)

data_renormalized_bacteria <- renorm_bacteria$data_renormalized
stats_subset_bacteria <- renorm_bacteria$stats_subset

# Pseudo-counts
absolute_counts_bacteria <-
  calculate_pseudo_counts(data_renormalized_bacteria, stats_subset_bacteria)

##FUNGI

# Prepare stats
stats_subset_table_fungi <- prepare_stats_table(
  stats_table = stats_table,
  metadata_general = metadata_smokers_general,
  metadata_target = metadata_smokers_decontam_fungi
)

stats_subset_fungi <- stats_subset_table_fungi$stats_subset
stats_subset_general_fungi <- stats_subset_table_fungi$stats_subset_general

# Renormalization
renorm_fungi <- renormalize_if_needed(
  data = data_species_filtered_decontam_fungi,
  stats_subset = stats_subset_fungi,
  read_column_suffix = "micop_fungi"
)

data_renormalized_fungi <- renorm_fungi$data_renormalized
stats_subset_fungi <- renorm_fungi$stats_subset

# Pseudo-counts
absolute_counts_fungi <-
  calculate_pseudo_counts(data_renormalized_fungi, stats_subset_fungi)


##VIRUS
data_species_filtered_virus <- filtered_virus$data_smokers

data_species_filtered_virus <- data_species_filtered_virus[rowSums(data_species_filtered_virus[,-1] != 0) > 1, ]


# Prepare stats
stats_subset_table_virus <- prepare_stats_table(
  stats_table = stats_table,
  metadata_general = metadata_smokers_general,
  metadata_target = metadata_smokers_virus
)

stats_subset_virus <- stats_subset_table_virus$stats_subset
stats_subset_general_virus <- stats_subset_table_virus$stats_subset_general

# Define read column suffix
read_column_suffix <- "micop_virus"

# -------------------------------
# 1. Target (decontam) virus renormalization
# -------------------------------
if (nrow(data_species_filtered_virus) > 20) {
  renorm_virus <- adjust_reads_and_normalize(
    data_species_filtered_virus,
    stats_subset_virus,
    read_column_suffix = read_column_suffix
  )
  
  data_renormalized_virus <- renorm_virus$renormalized_data
  stats_subset_virus <- renorm_virus$updated_stats
} else {
  data_renormalized_virus <- filtered_virus$data_smokers
  
  read_col <- paste0("Reads_after_dedup_", read_column_suffix)
  if (!(read_col %in% colnames(stats_subset_virus))) {
    stop(paste("Column", read_col, "not found in stats_subset_virus."))
  }
  
  stats_subset_virus$Reads_after_dedup_selected <- stats_subset_virus[[read_col]]
}

# Compute pseudo-counts
absolute_counts_virus <- calculate_pseudo_counts(data_renormalized_virus, stats_subset_virus)


```



## Post-decontamination Summary Statistics

Basic descriptive statistics are calculated for each species in the decontaminated dataset, including mean and median abundance, prevalence across samples, and absolute presence. These metrics provide an overview of species distribution and consistency across samples after decontamination and normalization.


```{r Stats }

compute_taxon_stats <- function(tax_data) {

  nsamples <- ncol(tax_data)

  tax_data$Mean <- apply(tax_data[, 2:nsamples], 1, mean)

  tax_data$Median <- apply(tax_data[, 2:nsamples], 1, median)

  tax_data$presence_proportion <- rowSums(tax_data[, 2:nsamples] > 0) / (nsamples - 1)

  tax_data$Abs_presence <- rowSums(tax_data[, 2:nsamples] > 0)

  return(tax_data)
}

```

```{r Stats }

##BACTERIA
taxData_decon_norm_stats_bacteria <-compute_taxon_stats(data_renormalized_bacteria)

# Optional: write metadata
#write.csv(metadata_smokers_decontam_bacteria,file = "../INPUT/metadata_bacteria_smokers_decontam.csv",row.names = FALSE)

##FUNGI
taxData_decon_norm_stats_fungi <- compute_taxon_stats(data_renormalized_fungi)

#write.csv(metadata_smokers_decontam_fungi,file = "../INPUT/metadata_fungi_smokers_decontam.csv",row.names = FALSE)


##VIRUS

taxData_decon_norm_stats_virus <- compute_taxon_stats(data_renormalized_virus)

#write.csv(metadata_smokers_decontam_virus,file = "../INPUT/metadata_virus_smokers_decontam.csv",row.names = FALSE)


```


## Descriptive Statistics of the Study Population

This section summarizes key demographic characteristics of the study population by smoking status, including sample size, sex distribution, and age statistics. A chi-squared test is performed to assess differences in sex distribution between groups, providing an overview of cohort comparability.

```{r Descriptive experimental setup}

library(gridExtra)
library(grid)

# Assuming metadata_smokers_labels is your dataframe
summary_table <- metadata_smokers %>%
  group_by(treatment) %>%  # Assuming 'treatment' indicates smoking status
  summarise(
    Total = n(),
    Female = sum(sex == "f", na.rm = TRUE),
    Male = sum(sex == "m", na.rm = TRUE),
    Age_Mean = mean(age, na.rm = TRUE),
    SD_Age = sd(age, na.rm = TRUE),
    Variance_Age = var(age, na.rm = TRUE)
  )

# Perform Chi-Squared Test on Sex Distribution

sex_table <- table(metadata_smokers$treatment, metadata_smokers$sex)
chi_test <- chisq.test(sex_table)

# Add Chi-Squared Test results
summary_table$Age_Mean <- round(summary_table$Age_Mean, 2)
summary_table$SD_Age <- round(summary_table$SD_Age, 2)
summary_table$Variance_Age <- round(summary_table$Variance_Age, 2)
summary_table$Chi_Squared_Test <- round(chi_test$statistic, 2)

colnames(summary_table) <-c("Treatment", "Total", "Female", "Male", "Mean (Age)", "SD (Age)", "Variance (Age)", "Chi-Squared Test (Sex)")

# Create the table without row names
table_grob <- tableGrob(summary_table, theme = ttheme_minimal(), rows = NULL)

# Add horizontal lines
lines <- gList(
  segmentsGrob(x0 = 0, x1 = 1, y0 = 0.98, y1 = 0.98, gp = gpar(lwd = 2)),  # Line below headers
  segmentsGrob(x0 = 0, x1 = 1, y0 = 0.02, y1 = 0.02, gp = gpar(lwd = 3))   # Bold line at bottom
)

# Save to PDF
#png("../../OUTPUT/Table_1.png",width = 8.5,height = 1.5,units = "in",res = 300)

#grid.draw(grobTree(table_grob, lines))

#dev.off()


```


## Prevalence Analysis

This section evaluates species prevalence across samples by summarizing the proportion of samples in which each species is detected. Prevalence distributions are visualized to characterize common and rare taxa in the decontaminated dataset prior to downstream comparative analyses.

```{r Analysis_1 Prevalesence}
source("Functions_presentation/Prevalence.R") 

### pre-analysis  ###

# Data_species only smokers cumulative plot

prevalence_bacteria <- count_prevalence_steps(taxData_decon_norm_stats_bacteria$presence_proportion)

# plot_prevalence_counts(prevalence_bacteria)

prevalence_fungi <- count_prevalence_steps(taxData_decon_norm_stats_fungi$presence_proportion)

# plot_prevalence_counts(prevalence_fungi)

prevalence_virus <- count_prevalence_steps(taxData_decon_norm_stats_virus$presence_proportion)

# plot_prevalence_counts(prevalence_virus)


```


## Taxonomic Composition and Relative Abundance

Species- and phylum-level relative abundances are calculated and visualized for the decontaminated, renormalized dataset. Low-abundance taxa are grouped to simplify interpretation, and mean relative abundances are computed per treatment group. Top taxa per smoking status are identified to highlight the most prevalent and abundant microbes in the cohort, providing a foundation for downstream comparative analyses.

```{r Analysis_2 Taxonomy Plot}
source("Functions_presentation/Taxonomy_relative_abundance.R") 
source("Functions_presentation/Taxonomy_RA_Mean.R") 
source("Functions_presentation/color_palettes.R") 

phyla_relative_abundance_mean_bacteria <- plot_relative_abundance_with_means(
  data_species = data_renormalized_bacteria,
  metadata = metadata_smokers_decontam_bacteria,
  level = "Phyla",
  low_abundance_threshold = 10,  # <-- here!
  colors_list = bacterial_phyla_colors,
  genus_colors = bacterial_phyla_colors, 
  group_var = "treatment",
  tag_label = "(A)"
)


names(bacterial_species_colors) <-gsub("_", " ", names(bacterial_species_colors))

# Example: Genus plot with 10% threshold for grouping low abundance taxa
species_relative_abundance_mean_bacteria <- plot_relative_abundance_with_means(
  data_species = data_renormalized_bacteria,
  metadata = metadata_smokers_decontam_bacteria,
  level = "Species",
  low_abundance_threshold = 10,  # <-- here!
  colors_list = bacterial_species_colors,
  genus_colors = bacterial_species_colors, 
  group_var = "treatment",
  tag_label = "(B)"
)


phylum_abundance_data_bacteria <- phyla_relative_abundance_mean_bacteria$abundance_data
species_abundance_data_bacteria <- species_relative_abundance_mean_bacteria$abundance_data


#Table with the most abundant bacterial microbes in the dataset--------------------------


# 3. Group and summarize: top N species per treatment
top_n <- 5  # change to how many top species you want per group

top_species_per_group_bacteria <- species_abundance_data_bacteria %>%
  group_by(treatment, Species) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))

#------------------------------------

# 2. Aggregate to phylum level per sample
library(dplyr)

phylum_abundance_cleaned_bacteria <- phylum_abundance_data_bacteria %>%
  select(-n, -label, -Color_Phyla, -Legend_Taxon) %>%
  group_by(samples, treatment, Phyla) %>%
  summarise(RA = round(sum(RA, na.rm = TRUE), 1), .groups = "drop")

# 5. Get top phyla per treatment
top_n <- 5
top_phyla_per_group_bacteria <- phylum_abundance_cleaned_bacteria %>%
  group_by(treatment, Phyla) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))


```

```{r Analysis_2 Taxonomy Plot}
# Load the function from the external R file
source("Functions_presentation/Taxonomy_relative_abundance.R") 
source("Functions_presentation/Taxonomy_RA_Mean.R") 
source("Functions_presentation/color_palettes.R") 


phyla_relative_abundance_mean_fungi <- plot_relative_abundance_with_means(
  data_species = data_renormalized_fungi,
  metadata = metadata_smokers_decontam_fungi,
  level = "Phyla",
  low_abundance_threshold = 10,  # <-- here!
  colors_list = fungi_phyla_colors,
  genus_colors = fungi_phyla_colors,
  group_var = "treatment",
  tag_label = "(C)"
)



# Genus plot with 10% threshold for grouping low abundance taxa


species_relative_abundance_mean_fungi <- plot_relative_abundance_with_means(
  data_species = data_renormalized_fungi,
  metadata = metadata_smokers_decontam_fungi,
  level = "Species",
  low_abundance_threshold = 10,  # <-- here!
  colors_list = fungi_species_colors,
  genus_colors = fungi_species_colors, 
  group_var = "treatment",
  tag_label = "(D)"
)



# Table with the most abundant bacterial microbes in the dataset--------------------------

phylum_abundance_data_fungi <- phyla_relative_abundance_mean_fungi$abundance_data
species_abundance_data_fungi <- species_relative_abundance_mean_fungi$abundance_data

# 3. Group and summarize: top N species per treatment
top_n <- 3  # change to how many top species you want per group

top_species_per_group_fungi <- species_abundance_data_fungi %>%
  group_by(treatment, Species) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))

#------------------------------------

# 2. Aggregate to phylum level per sample
library(dplyr)

phylum_abundance_cleaned_fungi <- phylum_abundance_data_fungi %>%
  select(-n, -label, -Color_Phyla, -Legend_Taxon) %>%
  group_by(samples, treatment, Phyla) %>%
  summarise(RA = round(sum(RA, na.rm = TRUE), 1), .groups = "drop")

# 5. Get top phyla per treatment
top_n <- 5
top_phyla_per_group_fungi <- phylum_abundance_cleaned_fungi %>%
  group_by(treatment, Phyla) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))

```

```{r Analysis_2 Taxonomy Plot}
# Load the function from the external R file
source("Functions_presentation/Taxonomy_relative_abundance_virus.R") 
source("Functions_presentation/color_palettes.R") 
source("Functions_presentation/Taxonomy_RA_Mean.R") 


order_relative_abundance_mean_virus <- plot_relative_abundance_with_means(
  data_species = data_renormalized_virus,
  metadata = metadata_smokers_virus ,
  level = "Order",
  low_abundance_threshold = 15,  # <-- here!
  colors_list = viral_order_colors,
  genus_colors = viral_order_colors,
  group_var = "treatment",
  tag_label = "(E)"
)



#  Genus plot with 10% threshold for grouping low abundance taxa


species_relative_abundance_mean_virus <- plot_relative_abundance_with_means(
  data_species = data_renormalized_virus,
  metadata = metadata_smokers_virus,
  level = "Species",
  low_abundance_threshold = 15,  # <-- here!
  colors_list = viral_species_colors,
  genus_colors = viral_species_colors, 
  group_var = "treatment",
  tag_label = "(F)"
)



# Table with the most abundant bacterial microbes in the dataset--------------------------

order_abundance_data_virus <- order_relative_abundance_mean_virus$abundance_data
species_abundance_data_virus <- species_relative_abundance_mean_virus$abundance_data

# 3. Group and summarize: top N species per treatment
top_n <- 5  # change to how many top species you want per group

top_species_per_group_virus <- species_abundance_data_virus %>%
  group_by(treatment, Species) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))

#------------------------------------

# 2. Aggregate to phylum level per sample
library(dplyr)

order_abundance_cleaned_virus <- order_abundance_data_virus %>%
  select(-n, -label, -Color_Order, -Legend_Taxon) %>%
  group_by(samples, treatment, Order) %>%
  summarise(RA = round(sum(RA, na.rm = TRUE), 1), .groups = "drop")

# 5. Get top phyla per treatment
top_n <- 5
top_order_per_group_virus <- order_abundance_cleaned_virus %>%
  group_by(treatment, Order) %>%
  summarise(mean_RA = mean(RA, na.rm = TRUE), .groups = "drop") %>%
  group_by(treatment) %>%
  slice_max(mean_RA, n = top_n, with_ties = FALSE) %>%
  arrange(treatment, desc(mean_RA))

```

## Alpha Diversity Analysis

Alpha diversity metrics are calculated to assess within-sample microbial diversity across smoking status groups. Here, the Shannon diversity index is computed for each sample, and statistical tests are performed to compare diversity between smokers and non-smokers, with visualizations generated for interpretation of group differences.

```{r Analysis_3 alpha diversity}

source("Functions_presentation/alpha_diversity_stars_pvalue.R") 
metrics <- c("diversity")  # can add "richness", "evenness", "simpson"

##BACTERIA
taxData_bacteria <- as.data.frame(data_renormalized_bacteria)
taxData_bacteria <- `rownames<-`(taxData_bacteria[, -1], taxData_bacteria[, 1])

for (m in metrics) {
  cat("Bacteria - metric:", m, "\n")
  
  result_alpha_bacteria <- alpha_diversity_analysis(
    species_df = taxData_bacteria,
    metadata_df = metadata_smokers_decontam_bacteria,
    metric = m,
    group_var = "treatment",
    group_levels = c("non-smoker", "smoker"),
    color_values = c("indianred1", "thistle4"),
    tag_label = "(A)", 
    y_expand = 1.6
  )
  
  print(result_alpha_bacteria$test)
  print(result_alpha_bacteria$normality)
}


##FUNGI
taxData_fungi <- as.data.frame(data_renormalized_fungi)
taxData_fungi <- `rownames<-`(taxData_fungi[, -1], taxData_fungi[, 1])

for (m in metrics) {
  cat("Fungi - metric:", m, "\n")
  
  result_alpha_fungi <- alpha_diversity_analysis(
    species_df = taxData_fungi,
    metadata_df = metadata_smokers_decontam_fungi,
    metric = m,
    group_var = "treatment",
    group_levels = c("non-smoker", "smoker"),
    color_values = c("indianred1", "thistle4"),
    tag_label = "(B)", 
    y_expand = 1.6
  )
  
  print(result_alpha_fungi$test)
  print(result_alpha_fungi$normality)
}


##VIRUS
taxData_virus <- as.data.frame(data_renormalized_virus)
taxData_virus <- `rownames<-`(taxData_virus[, -1], taxData_virus[, 1])

for (m in metrics) {
  cat("Virus - metric:", m, "\n")
  
  result_alpha_virus <- alpha_diversity_analysis(
    species_df = taxData_virus,
    metadata_df = metadata_smokers_virus,
    metric = m,
    group_var = "treatment",
    group_levels = c("non-smoker", "smoker"),
    color_values = c("indianred1", "thistle4"),
    tag_label = "(C)", 
    y_expand = 1.6
  )
  
  print(result_alpha_virus$test)
  print(result_alpha_virus$normality)
}

```


## Beta Diversity Analysis

Beta diversity is assessed to evaluate differences in overall microbial community composition between smokers and non-smokers. Principal coordinates analysis (PCoA) is performed using Bray–Curtis dissimilarities, and PERMANOVA is applied to test for statistically significant group differences. Visualizations include sample ordination plots with group-specific ellipses for interpretation of clustering patterns.

```{r Analysis_4 beta diversity}

#Load
source("Functions_presentation/beta_diversity_withpermanova.R") 

methods <- c("PCoA")  # Can add "PCA", "NMDS", "CAP"
custom_colors <- c("non-smoker" = "indianred1", "smoker" = "thistle4")

##BACTERIA

for (m in methods) {
  cat("Bacteria - beta diversity with method:", m, "\n")
  
  result_beta_bacteria <- beta_diversity_analysis(
    species_df = taxData_bacteria,
    metadata_df = metadata_smokers_decontam_bacteria, 
    method = m,
    color_var = "treatment",
    formula_vars = c("treatment"),
    color_values = custom_colors,
    ellipse = TRUE, 
    fill_label = "Smoking behaviour", 
    tag_label = "(D)"
  )
  
}

# Save outputs
beta_plot_bacteria <- result_beta_bacteria$plot
permanova_bacteria <- result_beta_bacteria$permanova
distance_matrix_bacteria <- result_beta_bacteria$distance_matrix

##FUNGI
for (m in methods) {
  cat("Fungi - beta diversity with method:", m, "\n")
  
  result_beta_fungi <- beta_diversity_analysis(
    species_df = taxData_fungi,
    metadata_df = metadata_smokers_decontam_fungi, 
    method = m,
    color_var = "treatment",
    formula_vars = c("treatment"),
    color_values = custom_colors,
    ellipse = TRUE, 
    fill_label = "Smoking behaviour", 
    tag_label = "(E)"
  )
}

beta_plot_fungi <- result_beta_fungi$plot
permanova_fungi <- result_beta_fungi$permanova
distance_matrix_fungi <- result_beta_fungi$distance_matrix


##VIRUS
for (m in methods) {
  cat("Virus - beta diversity with method:", m, "\n")
  
  result_beta_virus <- beta_diversity_analysis(
    species_df = taxData_virus,
    metadata_df = metadata_smokers_virus, 
    method = m,
    color_var = "treatment",
    formula_vars = c("treatment"),
    color_values = custom_colors,
    ellipse = TRUE, 
    fill_label = "Smoking behaviour", 
    tag_label = "(F)"
  )
}

beta_plot_virus <- result_beta_virus$plot
permanova_virus <- result_beta_virus$permanova
distance_matrix_virus <- result_beta_virus$distance_matrix


```


## Differential Abundance Testing with MaAsLin2

This step performs multivariable modeling to identify microbial taxa whose abundances are significantly associated with smoking status. Zero-inflated negative binomial models are applied to pseudo-count data, with fixed effects for treatment and random effects for batch, season, and sex. Multiple testing correction is applied to control false discovery, producing a list of taxa with significant associations for downstream interpretation.

```{r Analysis_5 Differential abundance with stats, eval=FALSE, include=FALSE}

# devtools::install_github("biobakery/maaslin2") # Install and load Maaslin2 if necessary
absolute_counts_tp <- t(absolute_counts)

rownames(metadata_smokers_decontam) <- metadata_smokers_decontam$sample_name
library(Maaslin2)

# Input: 
# dataset: a table of relative abundances (samples as rows, taxa as columns)
# metadata: metadata table with categorical or continuous variables of interest
metadata_smokers$Season_sample <- factor(metadata_smokers$Season_sample)
metadata_smokers$kit <- factor(metadata_smokers$kit)
rownames(metadata_smokers) <- metadata_smokers[,2]
# Running Maaslin2 with a fixed effect (e.g., "treatment") and random effect (e.g., "batch")
fit_data <- Maaslin2(
  input_data = absolute_counts_tp,          # Your relative abundance data
  input_metadata = metadata_smokers_decontam, # Your metadata table
  output = "../OUTPUT/Maaslin_output_pseudocounts_bacterial",            # Directory for output
  normalization = "NONE",               # No normalization
  transform = "NONE",                   # No transformation
  min_abundance = 0.01,
  min_prevalence = 0.35,
  analysis_method = "ZINB",               # Linear model
  max_significance = 0.05,              # Significance threshold
  fixed_effects =c("treatment"),     # Your fixed effect(s)
  random_effects = c("kit", "Season_sample", "sex"), # the difference you don't want to account for
  correction = "BH",                    # Benjamini-Hochberg correction for p-values
  standardize = FALSE,                  # No standardization
  cores = 1                             # Number of cores to use
)

```

## Comparative Analysis of Individual Species

This section visualizes the relative abundance of specific microbial species (e.g., Staphylococcus aureus) across smoking status groups using MaAsLin2 results. The approach highlights taxa with differential abundance and provides intuitive plots for interpretation of group-specific patterns.

```{r Comparative species analyisis}

library(readr)
# Load the function from the external R file
source("Functions_presentation/Comparative_test_Maaslin.R") 

result_A_bacteria <- plot_species_abundance(
  species_name = "Staphylococcus_aureus", #Moraxella_osloensis , Cutibacterium_acnes, Staphylococcus_aureus
  maaslin_result_file = "../OUTPUT/Maaslin_output_pseudocounts/all_results.tsv",
  abundance_data = absolute_counts_bacteria,
  metadata = metadata_smokers_decontam_bacteria, 
  treatment_col = "treatment", 
  custom_colors <- c("non-smoker" = "indianred1", "smoker" = "thistle4"),
  tag_label = "(A)" 
)

result_B_fungi <- plot_species_abundance(
  species_name = "Saccharomyces_cerevisiae", #Cryptococcus_neoformans , Saccharomyces_cerevisiae, Malassezia_globosa
  maaslin_result_file = "../OUTPUT/Maaslin_output_pseudocounts_fungal/all_results.tsv",
  abundance_data = absolute_counts_fungi,
  metadata = metadata_smokers_decontam_fungi, 
  treatment_col = "treatment", 
  colors = c("non-smoker" = "indianred1", "smoker" = "thistle4"),
  tag_label = "(B)" 
)

result_C_virus <- plot_species_abundance(
  species_name = "Glypta_fumiferanae_ichnovirus", #Cyprinid_herpesvirus_1 , Glypta_fumiferanae_ichnovirus, Pandoravirus_salinus
  maaslin_result_file = "../OUTPUT/Maaslin_output_pseudocounts_viral/all_results.tsv",
  abundance_data = absolute_counts_virus,
  metadata = metadata_smokers_virus, 
  treatment_col = "treatment", 
  colors = c("non-smoker" = "indianred1", "smoker" = "thistle4"),
  tag_label = "(C)"
)


```

```{r Composed_graphs_for_manuscript}


##Figure 1. Relative abundance of OSM in bacterial, fungal and viral kingdoms
library(ggplot2)
library(patchwork)

pA <- phyla_relative_abundance_mean_bacteria$plot
pB <- species_relative_abundance_mean_bacteria$plot

pC <- phyla_relative_abundance_mean_fungi$plot
pD <- species_relative_abundance_mean_fungi$plot

pE <- order_relative_abundance_mean_virus$plot
pF <- species_relative_abundance_mean_virus$plot


composed_figure <-(pA | pB) / (pC | pD) / (pE | pF)

# side by side


#ggsave("../OUTPUT/Figure_1.png",composed_figure, width = 32, height = 27, units = "in", dpi = 300)

##Figure 2.- alpha and beta diversity of OSM in bacterial, fungal and viral kingdoms

pdA <- result_alpha_bacteria$plot
pdB <- result_alpha_fungi$plot
pdC <- result_alpha_virus$plot

pdD <- result_beta_bacteria$plot
pdE <- result_beta_fungi$plot
pdF <- result_beta_virus$plot


composed_figure_2 <- (pdA | pdB | pdC) / (pdD | pdE | pdF)


#ggsave("../OUTPUT/Figure_2.png",composed_figure_2, width = 40, height = 20, units = "in", dpi = 300)


##Figure 3.- Differential expression analysis of bacterial, fungal and viral kingdoms

DAa <- result_A_bacteria$plot
DAb <- result_B_fungi$plot
DAc <- result_C_virus$plot

composed_figure_3 <- DAa | DAb | DAc


#ggsave("../OUTPUT/Supl_Figure_1.png",composed_figure_3, width = 30, height = 9, units = "in", dpi = 300)

```

