---
title: "Tear_Proteome_Analysis"
output: pdf_document
date: "2025-03-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Tear Proteome Analysis — Seasonal Influence

This R Markdown document is designed for the analysis of tear proteome data collected across different seasons. It integrates code, results, and documentation to provide a reproducible workflow for evaluating seasonal influences on protein expression.


```{r Library}

# Clean environment
rm(list = ls())

# Load required libraries
library(readxl)     # Reading Excel files
#install.packages("readxl")  # Install if not installed
library(tidyverse)  # Data wrangling and visualization

```

## Data Loading & Preparation
This section loads raw proteomic data, filters relevant datasets, and merges metadata to incorporate seasonal sampling information for downstream analysis.

```{r Data_loading, echo=FALSE}

# Check sheet names in the Excel file
sheet_names <- excel_sheets("../INPUT/20230717_SM_DZ_Smokers_humandatabase.xlsx")

# --- Load proteomic data ---

# Filtered protein groups used in differential expression testing
proteome_data_filt_hu <- read_excel(
  "../INPUT/20230717_SM_DZ_Smokers_humandatabase.xlsx",
  sheet = "proteinGroups DE test",
  col_types = "text")

# Experimental design for the samples
Experimental_design_hu <- read_excel(
  "../INPUT/20230717_SM_DZ_Smokers_humandatabase.xlsx",
  sheet = "Experimental design")

# Load metadata (includes seasonal information)
metadata <- read.csv("../INPUT/OSM_metadata_smokers_contact_complete.csv", sep = ",", header = TRUE)

```

##Data Preparation: Filter Quantified Protein Columns
This section selects columns corresponding to specific quantification methods and properly formatted sample identifiers from the filtered human proteomic dataset.

```{r Data_preparation}

# Copy filtered proteome data for transformation
proteome_data_hu_filtered <- proteome_data_filt_hu

# --- Define quantification methods of interest ---
quant_methods <- c("Top3", "MaxLFQ", "FragI", "iTop3", "iMaxLFQ", "iFragI")

# --- Create regex pattern to match quant method columns with valid sample IDs (e.g., 09_TPN) ---
# Pattern explanation:
#   - Any of the quant_methods
#   - Followed by digits and a sample code (e.g., _TPN)
quant_pattern <- paste0("(", paste(quant_methods, collapse = "|"), ").*\\d{2}_[A-Z]+(_[a-z])?")

# --- Extract matching column names based on the pattern ---
quant_cols <- grep(quant_pattern, names(proteome_data_hu_filtered), value = TRUE)

# --- Subset the data to retain gene info and quantification columns only ---
proteome_data_hu_filtered_trim <- proteome_data_hu_filtered %>%
  dplyr::select(Gene, Description, all_of(quant_cols))

```

##Data Transformation: iMaxLFQ Log2 Normalization
This section extracts the iMaxLFQ quantification data, performs a log₂ transformation, and removes any rows with missing values across samples.

```{r Data_preparation}

# --- Select columns corresponding to the "iMaxLFQ" quantification method ---
imaxlfq_cols <- grep("iMaxLFQ", colnames(proteome_data_hu_filtered_trim), value = TRUE)

# --- Subset the data to include only Gene and iMaxLFQ columns ---
proteome_imaxlfq <- proteome_data_hu_filtered_trim %>%
  dplyr::select(Gene, all_of(imaxlfq_cols))

# --- Define a log2 transformation function (with pseudocount to avoid log(0)) ---
log2_transform <- function(df, id_col = "Gene") {
  # Convert numeric columns, excluding identifier
  num_data <- df %>%
    dplyr::select(-all_of(id_col)) %>%
    mutate(across(everything(), as.numeric))
  
  # Apply log2 transformation with pseudocount
  log_data <- log2(num_data + 1)
  
  # Combine transformed data back with identifier
  bind_cols(df[[id_col]], log_data) %>%
    setNames(c(id_col, colnames(log_data)))}

# --- Apply transformation ---
imaxlfq_log2 <- log2_transform(proteome_imaxlfq)

# --- Remove rows with any NA values (complete cases only) ---
imaxlfq_clean <- imaxlfq_log2 %>%
  filter(if_all(-Gene, ~ !is.na(.)))

# Preview transformed data
head(imaxlfq_clean)

```

##Data Integration: Annotate iMaxLFQ Columns with Season Info
This step renames the iMaxLFQ columns to include seasonal context, based on the experiment ID, improving clarity for seasonal comparisons.


##Multivariate Analysis & Clustering: PCA, PERMANOVA, Heatmap
This section performs exploratory analysis to assess the influence of seasonal sampling on tear proteome profiles.

```{r Data_preparation}

expr_matrix <- as.matrix(imaxlfq_clean[,-1])
rownames(expr_matrix) <- imaxlfq_clean$Gene
colnames(expr_matrix) <- gsub("^iMaxLFQ\\s+", "", colnames(expr_matrix))

# Get unique sample names (columns)
sample_ids <- colnames(expr_matrix)

#Pad the numbers in the samplenames
Experimental_design_hu <- Experimental_design_hu %>%
  mutate(Sample.Name.Padded = str_replace(Sample.Name, "^([0-9])_", "0\\1_"))


# Subset metadata with relevant columns and unique rows
metadata <- Experimental_design_hu %>%
  select(Sample.Name.Padded, Nb.of.samples, Patientnumber, LongGroupName, Contains, Sex, Age, BMI) %>%
  distinct() %>%
  filter(Sample.Name.Padded %in% sample_ids)

#create a column treatment
metadata <- metadata %>%
  mutate(treatment = case_when(
    LongGroupName == "Smoker (Raucher)" ~ "smoker",
    LongGroupName == "Control (Normal)" ~ "non-smoker",
    TRUE ~ NA_character_  # in case there are other values
  ))


```

```{r PCA for general view}

# Load the function
source("Functions_presentation/beta_diversity_withpermanova.R") 

# Define the beta diversity methods to iterate through
methods <- c("PCA")

# Define colors
custom_colors <- c("non-smoker" = "indianred1", "smoker" = "thistle4")

for (m in methods) {
  cat("Processing beta diversity with method:", m, "\n")
  
  result_PCA <- beta_diversity_analysis(
    species_df = expr_matrix,
    metadata_df = metadata, 
    method = m,
    color_var = "treatment",
    #shape_var = "kit",
    formula_vars = c("treatment"),
    color_values = custom_colors,
    ellipse = T, 
    tag_label = "(A)"
  )
  
  # ⬇️ Save the PCA plot to a variable for later use
  if (m == "PCA") {
    pca_plot_object <- result_PCA$plot  # <-- This is your vector ggplot object
  }
}

```


##Diagnostic Plots: Expression Distribution & Sample Relationships
Visualize the expression data distribution per sample and relationships among samples to spot potential outliers or batch effects.

```{r continuation}

# Extract season group labels from column names
sample_names <- colnames(imaxlfq_clean)[-1]

smoking_group <- sub(".*_(TPR|TPN).*", "\\1", sample_names)
smoking_group <- factor(smoking_group)

library(limma)

# Boxplot: Expression values per sample (show distribution and variability)
boxplot(expr_matrix, 
        las = 2,                     # Rotate axis labels for readability
        main = "Expression values per sample",
        col = "lightblue")

# Density plot: visualize overall expression distribution per sample
plotDensities(expr_matrix, 
              col = rainbow(ncol(expr_matrix)),
              main = "Density Plot of Expression Values")

# Multidimensional scaling (MDS) plot: visualize sample similarity in 2D space
plotMDS(expr_matrix,
        col = as.numeric(smoking_group),  # Color points by season group
        pch = 16,                        # Solid circle points
        cex = 1.5,                       # Increase point size
        labels = NULL,                   # Hide sample labels for clarity
        main = "MDS Plot of Samples")

legend("topright", legend = levels(smoking_group), col = 1:length(levels(smoking_group)), pch = 16)

```


##Differential Expression Analysis & Venn Diagram Overlaps by Season

```{r New Analysis}

# Load necessary packages
library(limma)
library(VennDiagram)
library(grid)

# 1. Design matrix for season groups (no intercept)
design <- model.matrix(~0 + smoking_group)
colnames(design) <- levels(smoking_group)

# 2. Define pairwise contrasts between seasons
contrast_matrix <- makeContrasts(TPN - TPR, levels = design)  # Adjust if needed

# 3. Fit linear model and apply contrasts
fit <- lmFit(expr_matrix, design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# 4. Extract DE results for each pairwise contrast
results_TPN_vs_TPR <- topTable(fit2, number = Inf, adjust = "fdr")

# 5. Get significantly DE proteins (p-value < 0.05)
sig_prots_smoking <- results_TPN_vs_TPR$ID[results_TPN_vs_TPR$adj.P.Val < 0.05]

```

##Visualization: Volcano Plot for Smokers and non-Smokers Differential Expression


```{r Gene Visualization }

# Load EnhancedVolcano package for easy volcano plot visualization
library(EnhancedVolcano)

# Generate Volcano Plot for TPN vs TPR contrast
# Pick top 10 by p-value (or change to abs(logFC) if you prefer)
top10 <- results_TPN_vs_TPR %>%
  arrange(P.Value) %>%
  slice(1:10) %>%
  pull(ID)


volanoplot <- EnhancedVolcano(
  results_TPN_vs_TPR,
  lab = rep("", nrow(results_TPN_vs_TPR)),
  x = "logFC",
  y = "adj.P.Val",
  pCutoff = 0.05,
  FCcutoff = 1,
  col = c("grey", "gray60", "dodgerblue4", "firebrick1"),  # <-- custom colors
  pointSize = 4,  # adjust circle size
  labSize = 5,  # <-- increase/decrease this number to change label size
  ylim = c(0, 0.3),
  title = "Volcano Plot: smoker vs non-smoker",
  subtitle = "No FDR-significant proteins",
  caption = paste0("total = ", nrow(results_TPN_vs_TPR), " proteins")
) +
  labs(tag = "(B)") +   # ✅ panel label
  theme(
    axis.text.x = element_text(size = 22),
    axis.text.y = element_text(size = 22),
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 22),
    legend.text = element_text(size = 35),
    legend.title = element_text(size = 35),
    legend.position = "bottom",
    #plot.title = element_text(size = 25, face = "bold"),
    plot.title = element_blank(),
    #plot.subtitle = element_text(size = 18, , face = "bold"),
    plot.subtitle = element_blank(),
    plot.caption = element_blank(), 
    plot.tag = element_text(face = "bold", size = 30),
    plot.tag.position = c(0, 1.05)
    )

```

```{r Figure_Composition }

library(ggplot2)
library(patchwork)

pA <- result_PCA$plot
pB <- volanoplot

composed_figure <- pA | pB

# side by side

#ggsave("../OUTPUT/Figure_3.png",composed_figure, width = 30, height = 9, units = "in", dpi = 300)


```


